---
title: "DS4002-Project1"
author: "Anna Stein"
date: "9/12/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---



```{r}
install.packages("readxl")
library("readxl")
annual_movement_data = read_excel("/Users/Anna/Downloads/hst_mig_a_1 (3).xlsx")
annual_movement_data = na.omit(annual_movement_data)
```


changing column headings
```{r}
library(dplyr)
names(annual_movement_data)[2] = "Total Population (over 1 year old)"
names(annual_movement_data)[3] = "Total Non-Movers"
names(annual_movement_data)[4] = "Total Movers"
names(annual_movement_data)[5] = "REMOVE: Total Movers" #actually, just take this column out ?? 
names(annual_movement_data)[6] = "Total Movers that stayed in same county"
names(annual_movement_data)[7] = "Total Movers that moved to different county"
names(annual_movement_data)[8] = "Total Movers: different county, same state"
names(annual_movement_data)[9] = "Total Movers: different county, different state"

# need to take out column 5 and column 10 

drop = c("REMOVE: Total Movers", "...10")
annual_mvmt_data = annual_movement_data[,!(names(annual_movement_data)%in%drop)]
# going forward new dataset 

names(annual_mvmt_data)[1] = "Mobility Period"
```

going to drop rows that denote the percentages instead of the levels (just personal preference)
```{r}
annual_mvmt_data = annual_mvmt_data[-c(72:142),]
```

exploratory questions: 
- what proportion of the total U.S. population moves each year? 
- out of total movers, what proportion are staying in the same state (same or different county) vs. leaving the state? 

can create new columns for each of these proportions 

```{r}
#annual_mvmt_data$ProportionOfMovers = annual_mvmt_data$`Total Movers` / annual_mvmt_data$`Total Population (over 1 year old)` # didn't work bc columns are characters, need to be numeric

i = c(2:8)
annual_mvmt_data[,i] = apply(annual_mvmt_data[,i], 2, function(x) as.numeric(as.character(x)))
# now columns are numeric 

annual_mvmt_data$ProportionOfMovers = annual_mvmt_data$`Total Movers` / annual_mvmt_data$`Total Population (over 1 year old)`
```

```{r}
annual_mvmt_data$`Mobility Period` = as.factor(annual_mvmt_data$`Mobility Period`)
# want to only have one year as the mobility period, the second year listed
annual_mvmt_data$`Mobility Period` = substr(annual_mvmt_data$`Mobility Period`, 6, 9)
```

Line Graph 1 
```{r}
library(ggplot2)
line_graph = ggplot(data=annual_mvmt_data, aes(x=annual_mvmt_data$`Mobility Period`,y=annual_mvmt_data$ProportionOfMovers, group=1)) + geom_line(color="#aa0022",size=1)+scale_x_discrete(breaks=c("1950","1960","1970","1980","1990","2000", "2010", "2020")) + labs(x="Year", y="Proportion of Total Population that Moved") + ylim(0.000,0.300)

line_graph
```


Now want to answer the question: out of total movers, what proportion are staying in the same state (same or different county) vs. leaving the state? 
```{r}
# make a new column for movers that are staying in the same state
# make a new column for movers that are leaving the state

annual_mvmt_data$OutOfStateMovers = annual_mvmt_data$`Total Movers: different county, different state`/annual_mvmt_data$`Total Movers`

annual_mvmt_data$WithinStateMovers = (annual_mvmt_data$`Total Movers that stayed in same county` + annual_mvmt_data$`Total Movers: different county, same state`)/annual_mvmt_data$`Total Movers`
```

can make a line graph showing how the proportion of total movers that leave their home state
has changed over time 
```{r}
line_graph1 = ggplot(data=annual_mvmt_data, aes(x=annual_mvmt_data$`Mobility Period`,y=annual_mvmt_data$OutOfStateMovers, group=1)) + geom_line(color="#aa0023",size=1)+scale_x_discrete(breaks=c("1950","1960","1970","1980","1990","2000", "2010", "2020")) + labs(x="Year", y="Proportion of Total Movers that Moved to a Different State") + ylim(0.05,0.30)

line_graph1
```
This chart isn't that interesting  ... seems like the proportion of people that move that leave for a different state hasn't changed a ton over time (since roughly 1950)

Conclusions from first dataset: 
1. The rate at which people move has declined over time ... the proportion of 'movers' for each year relative to the whole population has trended down overall since ~1950. The beginning of the faster decrease in moves began around the 1980s 













Going to move on to dataset that includes reason for moving
```{r}
reasons_data = read_excel("/Users/Anna/Downloads/hst_mig_a_5 (3).xlsx")
```


renaming columns
```{r}
names(reasons_data)[1] = "Mobility Period"
names(reasons_data)[2] = "Total Movers (Over 1 yrs old)"
names(reasons_data)[3] = "Family: Change in Marital Status"
names(reasons_data)[4] = "Family: To Establish Own Household"
names(reasons_data)[5] = "Family: Other"
names(reasons_data)[6] = "Employment: New Job or Job Transfer"
names(reasons_data)[7] = "Employment: Look for Work, or Lost Job"
names(reasons_data)[8] = "Employment: To be Closer to Work"
names(reasons_data)[9] = "Employment: Retired"
names(reasons_data)[10] = "Employment: Other"
names(reasons_data)[11] = "Housing: Wanted to Own, Not Rent Home"
names(reasons_data)[12] = "Housing: Wanted Newer/Better/Larger Home"
names(reasons_data)[13] = "Housing: Wanted Better/Safer Neighborhood"
names(reasons_data)[14] = "Housing: Wanted Cheaper Housing"
names(reasons_data)[15] = "Housing: Foreclosure/Evicted"
names(reasons_data)[16] = "Housing: Other"
names(reasons_data)[17] = "Misc: Relationship with Unmarried Partner"
names(reasons_data)[18] = "Misc: to Attend or Leave College"
names(reasons_data)[19] = "Misc: Change in Climate"
names(reasons_data)[20] = "Misc: Health Reasons"
names(reasons_data)[21] = "Misc: Natural Disaster"
names(reasons_data)[22] = "Misc: Other"
```

```{r}
# now remove the next couple rows that held the names 
reasons_data = reasons_data[6:89, ]
#View(reasons_data)
```

```{r}
# want 'mobility period' to focus on the second year listed rather than the range 
reasons_data$`Mobility Period` = substr(reasons_data$`Mobility Period`, 6, 9)
```

```{r}
reasons_data = na.omit(reasons_data)
```

```{r}
# going to take out percentage rows/columns in favor of the row values that are numbers of people
reasons_data = reasons_data[1:25,]
```

```{r}
# going to make columns for the total of each major category: Family, Employment, Housing, Misc.
# first need to make columns numeric
i = c(2:22)
reasons_data[,i] = apply(reasons_data[,i], 2, function(x) as.numeric(as.character(x)))

total_family = reasons_data$`Family: Change in Marital Status` + reasons_data$`Family: To Establish Own Household` + reasons_data$`Family: Other`
reasons_data$`Total Family-Related Moves` = total_family 

total_emp = reasons_data$`Employment: New Job or Job Transfer` + reasons_data$`Employment: Look for Work, or Lost Job`+reasons_data$`Employment: To be Closer to Work` + reasons_data$`Employment: Retired`+reasons_data$`Employment: Other`
reasons_data$`Total Employment-Related Moves` = total_emp



housing1 = reasons_data$`Housing: Wanted to Own, Not Rent Home` + reasons_data$`Housing: Wanted Newer/Better/Larger Home`
housing2 = reasons_data$`Housing: Wanted Better/Safer Neighborhood` + reasons_data$`Housing: Wanted Cheaper Housing`
housing3 = reasons_data$`Housing: Other`
reasons_data$total_housing = housing1 + housing2 + housing3

#total_housing = reasons_data$`Housing: Wanted to Own, Not Rent Home` + reasons_data$`Housing: Wanted Newer/Better/Larger Home` + reasons_data$`Housing: Wanted Better/Safer Neighborhood` + reasons_data$`Housing: Wanted Cheaper Housing` + reasons_data$`Housing: Foreclosure/Evicted` + reasons_data$`Housing: Other`
#reasons_data$`Total Housing-Related Moves` = total_housing

# now we have 3 new columns for the total number of moves that fall into 3 of the main categories (family, employment, and housing)

# now, going to make 3 more columns that give the proportion of total moves that were due to the overall categories of housing, family, or employment

reasons_data[is.na(reasons_data)] = 0 # replacing NAs with zeros so that we can still get proportions 
# idk if this is a legit strategy but it should work for now
# because we aren't doing averages or anyting like that yet 


reasons_data$`Proportion of Total Moves that were Family-Related` = reasons_data$`Total Family-Related Moves`/reasons_data$`Total Movers (Over 1 yrs old)`

reasons_data$`Proportion of Total Moves that were Employmt-Related` = reasons_data$`Total Employment-Related Moves`/reasons_data$`Total Movers (Over 1 yrs old)`

reasons_data$`Proportion of Total Moves that were Housing-Related` = reasons_data$total_housing/reasons_data$`Total Movers (Over 1 yrs old)`


## something is wrong becasue it is seeing the 0s in certain columns (like for one of the housing columns) and making the 'total housing' number also = 0
# maybe need to convert these specific columns back to numeric??
reasons_data$`Housing: Foreclosure/Evicted` = as.numeric(reasons_data$`Housing: Foreclosure/Evicted`)

reasons_data$`Misc: Relationship with Unmarried Partner` = as.numeric(reasons_data$`Misc: Relationship with Unmarried Partner`)
# that didn't change the problem, will have to revisit this
```


## Line graphs:
1. Over the past 20 years, what proportion of moves were housing related in some way (each year)? 
```{r}
line_graph2 = ggplot(data=reasons_data, aes(x=reasons_data$`Mobility Period`,y=reasons_data$`Proportion of Total Moves that were Housing-Related`, group=1)) + geom_line(color="darkblue",size=1)+ labs(x="Year", y="Proportion of Total Moves that were Housing-Related")+ylim(0.0,0.6)+scale_x_discrete(breaks=c("1990", "2000", "2010", "2020"))+geom_hline(yintercept = mean(reasons_data$`Proportion of Total Moves that were Housing-Related`), color="cornflowerblue")

line_graph2
```


2. Over the past 20 years, what proportion of moves were employment related in some way (each year)?

```{r}
line_graph3 = ggplot(data=reasons_data, aes(x=reasons_data$`Mobility Period`,y=reasons_data$`Proportion of Total Moves that were Employmt-Related`, group=1)) + geom_line(color="chartreuse4",size=1)+ labs(x="Year", y="Proportion of Total Moves that were Emplymt-Related")+ylim(0.0,0.6)+scale_x_discrete(breaks=c("1990", "2000", "2010", "2020"))+geom_hline(yintercept = mean(reasons_data$`Proportion of Total Moves that were Employmt-Related`), color="chartreuse3")

line_graph3
```


Over the past 20 years, what proportion of moves were family-related in some way (each year)? 
```{r}
line_graph4 = ggplot(data=reasons_data, aes(x=reasons_data$`Mobility Period`,y=reasons_data$`Proportion of Total Moves that were Family-Related`, group=1)) + geom_line(color="mediumorchid4",size=1)+ labs(x="Year", y="Proportion of Total Moves that were Family-Related")+ylim(0.0,0.6)+scale_x_discrete(breaks=c("1990", "2000", "2010", "2020"))+geom_hline(yintercept = mean(reasons_data$`Proportion of Total Moves that were Family-Related`), color="mediumorchid1")

line_graph4
```

writing new dataframe into a CSV
```{r}
write.csv(reasons_data, "C://Users/Anna/DS3001/DS-3001-updated/reasons_data.csv", row.names=TRUE)
```

```{r}
# need to take out the column for foreclosure/eviction due to high amount of missing values 
reasons_data = reasons_data[,-c(15)]
```

```{r}
# need to take out the column called "total housing-related moves" because its redundant, just want to use "total_housing" column
reasons_data = reasons_data[,-c(24)]
```




